\textbf{Two-point statistics without bins: A continuous-function generalization of the correlation function estimator for large-scale structure}

### 2020-05-05 paper/paper.tex
paper tweaks
hogg responses to ksf comments

% language
\newcommand{\cf}{2pcf\xspace} %2pF? 2PCF? %TODO: fix spacing after
% to capitalize or not to capitalize? leaning no
\newcommand{\Est}{The Continuous-Function Estimator\xspace}
\newcommand{\est}{the Continuous-Function Estimator\xspace}
\newcommand{\LS}{LS\xspace}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\etc}{\foreign{etc}}
% math
\newcommand{\inv}{^{-1}}
\newcommand{\T}{^{\mathsf{T}}}
\newcommand{\hmpc}{$h^{-1}$Mpc}
% comments
\newcommand{\KSF}[1]{\textcolor{teal}{KSF says: #1}}
%\newcommand{\KSF}[1]{\textcolor{teal}{}}

@@

\begin{abstract}\noindent
% Context
The two-point correlation function (\cf) is the most important statistic in structure formation, used to measure the clustering of density field tracers (e.g. galaxies).
Current estimators of the 2pcf, including the standard Landy-Szalay (\LS) estimator, have significant limitations.
%Aims
In this work we address the issue that the \LS estimator evaluates the \cf in bins of separation between objects, which results in a loss of information and a trade-off between bias and variance. \KSF{there are other aims but they are hard to describe, as they are not ``issues'' with the LS estimator but other capabilities of ours - how to phrase?}
%Methods
We present a new estimator for the \cf, \emph{\est}, which generalizes \LS to a continuous representation and obviates binning in separation or any other property. 
Our estimator replaces the binned pair counts of \LS with a linear superposition of any set of basis functions, and outputs the best-fit linear combination of basis functions to describe the \cf. 
It is closely related to the information-theory optimal estimator used in linear least-squares fitting. The choice of basis can take into account the expected form of the \cf, as well as its dependence on other properties beyond separation.
%Results
We show that \est can estimate the clustering of artificial data in representations that provide more accuracy with fewer basis functions than \LS.
Using a spline basis representation, we show that \est achieves lower bias and lower variance than \LS. 
We also demonstrate how the estimator can be used to directly estimate the Baryon Acoustic Oscillation scale.
Critically, these representations reduce the number of mock catalogs required for covariance estimation, a limiting factor in \cf measurements.
%As a demonstration, we apply the estimator to the Luminous Red Galaxy sample of the Sloan Digital Sky Survey (SDSS). 
%We show that the continuous-function estimator, with basis functions chosen to be derivatives with respect to the cosmological parameters, provides a direct measurement of cosmological parameters from galaxy observations.
%Similarly, we use the estimator to directly measure the location of the baryon acoustic feature in the SDSS data.
We discuss other applications and limitations of \est for present and future studies of large-scale structure, including determining the dependence of clustering on galaxy properties and potentially unifying real-space and Fourier-space approaches to clustering measurements.

@@

The large-scale structure (LSS) of the universe is a critical probe of fundamental cosmology. 
It encodes information about the physics of the early universe and the subsequent expansion history.
In particular, the LSS provides a probe of the Baryon Acoustic Oscillations (BAO), density fluctuations resulting from baryon-photon coupling in the early universe.
The distance traveled by these waves imprints a feature on the statistical description of the LSS, which can be used to determine the characteristic BAO length scale \citep{EisensteinHu1998}.
The LSS also contains the signature of redshift-space distortions caused by the peculiar velocities of galaxies, which probe the growth rate of structure \citep{Kaiser1987}.
Additionally, the LSS can be used to constrain galaxy formation in conjunction with models of galaxy bias (e.g. \citealt{Hamilton1988}). %?
With current observations, the LSS is well-described by a cold dark matter model with a cosmological constant, the standard $\Lambda$CDM model.
Upcoming galaxy surveys will observe larger volumes with improved measurements, allowing us to test $\Lambda$CDM to even higher precision.
\KSF{Should I name-drop surveys, and their data volume / sky area?}

We characterize the LSS by using luminous sources to trace the underlying matter density field.
These tracers are often to taken to be galaxies, but can also be galaxy clusters, quasars and other sources.
\KSF{should i say that hereafter we will take them to be galaxies?}
The clustering of these objects is measured with two-point statistics, namely the power spectrum $P(k)$ and the two-point correlation function (\cf).
These characterize the clustering in Fourier space and real space, respectively, with the \cf defined as the Fourier Transform of the power spectrum: 
\begin{equation}
\xi(r) = \frac{1}{2\pi^2} \int_0^{\infty} P(k) \frac{\mathrm{sin}(kr)}{kr} k^2 dk
\end{equation}
where we assume isotropy. \KSF{should i give the original Fourier transform before itegrating over orientation?}
In principle, the power spectrum and the two-point correlation function contain the same information.
However, in practical applications the survey boundaries introduce nontrivial issues in computing these statistics, leading to diverging approaches to their computation with a significant difference in expense.
The \cf requires more computational power and extra survey products, but it is an incredibly useful tool; for instance, it well-suited to the analysis of the BAO feature which manifests at a single scale in real space.

@@

We assume we have a data catalog with $N_D$ objects and a random catalog with $N_R$ objects.
The pair counts for \LS and related estimators are then defined explicitly as
%TODO: fix spacing on sides of equal sign
\begin{eqnarray}\displaystyle
DD_k &\equiv& \frac{2}{N_D(N_D-1)} \sum_{n n'} i(g_k < |x_n - x_{n'}| < h_k) \\
DR_k &\equiv& \frac{1}{N_D N_R} \sum_{n m} i(g_k < |x_n - x_m| < h_k) \\
RR_k &\equiv& \frac{2}{N_R(N_R-1)} \sum_{m m'} i(g_k < |x_m - x_{m'}| < h_k),
\end{eqnarray}
where $DD_k$ is the count of data--data tracer pairs in bin $k$ (which has bin edges $g_k$ and $h_k$), $i()$ is an indicator function \KSF{define?}, $x$ is the tracer position, the $n$ and $n'$ indices index data positions, the $m$ and $m'$ indices index random catalog positions, $DR_k$ is the count of data--random pairs, and $RR_k$ is the count of random--random pairs.
\KSF{Do i need to note that Corrfunc (and some papers) actually kills the -1 in the prefactor? though this might be related to how Corrfunc does pair counting}

@@

\KSF{Figure demonstrating trade-off between bias and variance? or too simple}

\KSF{Should i explain / show some of the other estimators? --> now i do in the beyond LS section}

\KSF{Do i need to explain real space vs redshift space vs projected correlation function?}

@@

We generalize the \LS estimator defined above to any set of $K$ basis functions $f$ of the pair, so we now have
\begin{eqnarray}\displaystyle
DD &\equiv& \frac{2}{N_D(N_D-1)} \sum_{n n'} f(T_n, T_{n'}) \\
DR &\equiv& \frac{1}{N_D N_R} \sum_{n m} f(T_n, T_{m}) \\
RR &\equiv& \frac{2}{N_R(N_R-1)} \sum_{m m'} f(T_m, T_{m'}) \\
QQ &\equiv& \frac{2}{N_R(N_R-1)} \sum_{m m'} f(T_m, T_{m'}) \cdot f\T(T_m, T_{m'}),
\end{eqnarray}
%introduced this T notation to be the most general but it's a bit clunky (see below)
%what do we think about K-vector notation?
%matrix->tensor?
where $DD$, $DR$, and $RR$ are now $K$-vectors, $T$ refers to the data for the given tracer, and $QQ$ is a $K$-by-$K$ matrix defined as the outer product of basis function evaluations of the random--random pairs.

@@

\KSF{Worth giving analytic formulation of RR term? Or at least mentioning that there is one and we have derived it?}

\KSF{Discuss the limit of infinitesimal bins? Do we know what this is?}

We implement this estimator based on the correlation function package \texttt{Corrfunc} by \cite{Sinha2019}.
The code is open-source and available at \texttt{github.com/kstoreyf/Corrfunc}.
\KSF{Should give any detail about implementation?}

\KSF{where to put this link? In a footnote? And is it ok to keep it as a fork?}

@@

\KSF{Does this count as Applications if its on fake data? Better name?}

@@

\label{fig:spline}
\begin{figure}[ht]
\centering
    \includegraphics[width=0.8\textwidth]{tophat_cubic}
    \caption{A comparison between \est with a cubic spline basis function (red dotted) and the standard tophat estimator (blue solid). The shaded region is the 1$\sigma$ variation in the 100 mock catalogs. The cubic spline estimator has lower bias and lower variance with fewer bins. \KSF{quote rmse in caption? otherwise not clear from figure its less biased.} \KSF{does tophat need to be a different linestyle?}} \KSF{fix to dotted linestyle}

@@

\KSF{Should i have a figure showing the basis functions?}

>>>>

% language
\newcommand{\cf}{2pcf\xspace} %2pF? 2PCF? %TODO: fix spacing after
% to capitalize or not to capitalize? leaning no
\newcommand{\Est}{The Continuous-Function Estimator\xspace}
\newcommand{\est}{the Continuous-Function Estimator\xspace}
\newcommand{\LS}{LS\xspace}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\etc}{\foreign{etc}}
% math
\newcommand{\inv}{^{-1}}
\newcommand{\T}{^{\mathsf{T}}}
\newcommand{\hmpc}{$h^{-1}$Mpc}
% comments
\newcommand{\KSF}[1]{\textcolor{teal}{KSF says: #1}}
\newcommand{\hogg}[1]{\textcolor{teal}{Hogg says: #1}}
%\newcommand{\KSF}[1]{\textcolor{teal}{}}

@@

\begin{abstract}\noindent
% Context
The two-point correlation function (\cf) is the most important statistic in structure formation, used to measure the clustering of density field tracers (e.g. galaxies).
Current estimators of the 2pcf, including the standard Landy-Szalay (\LS) estimator, have significant limitations.
%Aims
In this work we address the issue that the \LS estimator evaluates the \cf in bins of separation between objects, which results in a loss of information and a poor trade-off between bias and variance, given the inappropriateness of hard bin edges in most scientific contexts. \hogg{The issue is NOT the bias--variance trade-off; that exists everywhere. It is that this trade-off is bad when the model is inappropriate. It is the inappropriateness that we are addressing. I still don't like this wording, but I tried.}
%Methods
We present a new estimator for the \cf, \emph{\est}, which generalizes \LS to a continuous representation and obviates binning in separation or any other property. 
Our estimator replaces the binned pair counts of \LS with a linear superposition of any set of basis functions, and outputs the best-fit linear combination of basis functions to describe the \cf. 
It is closely related to the information-theory optimal estimator used in linear least-squares fitting. The choice of basis can take into account the expected form of the \cf, as well as its dependence on other properties beyond separation.
%Results
We show that \est can estimate the clustering of artificial data in representations that provide more accuracy with fewer basis functions than \LS.
Using a spline basis representation, we show that \est achieves lower bias and lower variance than \LS. 
We also demonstrate how the estimator can be used to directly estimate the Baryon Acoustic Oscillation scale.
Critically, these representations reduce the number of mock catalogs required for covariance estimation, a limiting factor in \cf measurements.
%As a demonstration, we apply the estimator to the Luminous Red Galaxy sample of the Sloan Digital Sky Survey (SDSS). 
%We show that the continuous-function estimator, with basis functions chosen to be derivatives with respect to the cosmological parameters, provides a direct measurement of cosmological parameters from galaxy observations.
%Similarly, we use the estimator to directly measure the location of the baryon acoustic feature in the SDSS data.
We discuss other applications and limitations of \est for present and future studies of large-scale structure, including determining the dependence of clustering on galaxy properties and potentially unifying real-space and Fourier-space approaches to clustering measurements.

@@

The large-scale structure (LSS) of the universe is a critical probe of fundamental cosmology. 
It encodes information about the physics of the early universe and the subsequent expansion history.
In particular, the LSS provides a probe of the Baryon Acoustic Oscillations (BAO), density fluctuations resulting from baryon-photon coupling in the early universe.
The distance traveled by these waves imprints a feature on the statistical description of the LSS, which can be used to determine the characteristic BAO length scale \citep{EisensteinHu1998}.
The LSS also contains the signature of redshift-space distortions caused by the peculiar velocities of galaxies, which probe the growth rate of structure \citep{Kaiser1987}.
Additionally, the LSS can be used to constrain galaxy formation in conjunction with models of galaxy bias (e.g. \citealt{Hamilton1988}). %?
With current observations, the LSS is well-described by a cold dark matter model with a cosmological constant, the standard $\Lambda$CDM model.
Upcoming galaxy surveys will observe larger volumes with improved measurements, allowing us to test $\Lambda$CDM to even higher precision.
\KSF{Should I name-drop surveys, and their data volume / sky area?} \hogg{I don't think it is necessary; use your judgement.}

We characterize the LSS by using luminous sources to trace the underlying matter density field.
These tracers are often to taken to be galaxies, but can also be galaxy clusters, quasars and other sources.
\KSF{should i say that hereafter we will take them to be galaxies?} \hogg{Yes.}
The clustering of these objects is measured with two-point statistics, namely the power spectrum $P(k)$ and the two-point correlation function (\cf).
These characterize the clustering in Fourier space and real space, respectively, with the \cf defined as the Fourier Transform of the power spectrum: 
\begin{equation}
\xi(r) = \frac{1}{2\pi^2} \int_0^{\infty} P(k) \frac{\mathrm{sin}(kr)}{kr} k^2 dk
\end{equation}
where we assume isotropy. \KSF{should i give the original Fourier transform before itegrating over orientation?} \hogg{Yes, and say that the second follows from the first.}
In principle, the power spectrum and the two-point correlation function contain the same information.
However, in practical applications the survey boundaries introduce nontrivial issues in computing these statistics, leading to diverging approaches to their computation with a significant difference in expense.
The \cf requires more computational power and extra survey products, but it is an incredibly useful tool; for instance, it well-suited to the analysis of the BAO feature which manifests at a single scale in real space.

@@

We assume we have a data catalog with $N_D$ objects and a random catalog with $N_R$ objects.
The pair counts for \LS and related estimators are then defined explicitly as
%TODO: fix spacing on sides of equal sign
\begin{eqnarray}\displaystyle
DD_k &\equiv& \frac{2}{N_D(N_D-1)} \sum_{n n'} i(g_k < |x_n - x_{n'}| < h_k) \\
DR_k &\equiv& \frac{1}{N_D N_R} \sum_{n m} i(g_k < |x_n - x_m| < h_k) \\
RR_k &\equiv& \frac{2}{N_R(N_R-1)} \sum_{m m'} i(g_k < |x_m - x_{m'}| < h_k),
\end{eqnarray}
where $DD_k$ is the count of data--data tracer pairs in bin $k$ (which has bin edges $g_k$ and $h_k$), $i()$ is an indicator function \KSF{define?} \hogg{yes}, $x$ is the tracer position, the $n$ and $n'$ indices index data positions, the $m$ and $m'$ indices index random catalog positions, $DR_k$ is the count of data--random pairs, and $RR_k$ is the count of random--random pairs.
\KSF{Do i need to note that Corrfunc (and some papers) actually kills the -1 in the prefactor? though this might be related to how Corrfunc does pair counting} \hogg{I have no idea what you are asking here.}

@@

\KSF{Figure demonstrating trade-off between bias and variance? or too simple} \hogg{Only if the point isn't clear without it.}

\KSF{Should i explain / show some of the other estimators? --> now i do in the beyond LS section} \hogg{No; our job is not to work through all estimators. But we SHOULD note that our mod can be applied to almost any estimator that currently uses bins!}

\KSF{Do i need to explain real space vs redshift space vs projected correlation function?} \hogg{No, but always be absolutely unambiguous about what space you are in. To the point that you might even want to use different symbols.}

@@

% there's a reason im not using k here! because these need to be vectors / tensors to then compute the amplitude
We generalize the \LS estimator defined above to any set of $K$ basis functions $f$ of the pair, so we now have
\begin{eqnarray}\displaystyle
DD &\equiv& \frac{2}{N_D(N_D-1)} \sum_{n n'} f(T_n, T_{n'}) \\
DR &\equiv& \frac{1}{N_D N_R} \sum_{n m} f(T_n, T_{m}) \\
RR &\equiv& \frac{2}{N_R(N_R-1)} \sum_{m m'} f(T_m, T_{m'}) \\
QQ &\equiv& \frac{2}{N_R(N_R-1)} \sum_{m m'} f(T_m, T_{m'}) \cdot f\T(T_m, T_{m'}),
\end{eqnarray}
%introduced this T notation to be the most general but it's a bit clunky (see below)
%what do we think about K-vector notation?
%matrix->tensor?
where $DD$, $DR$, and $RR$ are now $K$-vectors, $T$ refers to the data for the given tracer, and $QQ$ is a $K$-by-$K$ matrix defined as the outer product of basis function evaluations of the random--random pairs.

@@

\KSF{Worth giving analytic formulation of RR term? Or at least mentioning that there is one and we have derived it?} \hogg{Huh? We have one?}

\KSF{Discuss the limit of infinitesimal bins? Do we know what this is?} \hogg{Yes, we should discuss this, and show how we are related to that.}

We implement this estimator based on the correlation function package \texttt{Corrfunc} by \cite{Sinha2019}.
The code is open-source and available at \texttt{github.com/kstoreyf/Corrfunc}.
\KSF{Should give any detail about implementation?} \hogg{Yes, we should summarize why this is a good package to build on.}

\KSF{where to put this link? In a footnote? And is it ok to keep it as a fork?} \hogg{Put link here and in a note at the end of the acks at the end which says ``all of the code is available open-source at...}

@@

\KSF{Does this count as Applications if its on fake data? Better name?} \hogg{I say ``Experiments and results'' these days.}

@@

\label{fig:spline}
\begin{figure}[ht]
\centering
    \includegraphics[width=0.8\textwidth]{tophat_cubic}
    \caption{A comparison between \est with a cubic spline basis function (red dotted) and the standard tophat estimator (blue solid). The shaded region is the 1$\sigma$ variation in the 100 mock catalogs. The cubic spline estimator has lower bias and lower variance with fewer bins. \KSF{quote rmse in caption? otherwise not clear from figure its less biased.} \hogg{yes} \KSF{does tophat need to be a different linestyle?}} \KSF{fix to dotted linestyle}

@@

\KSF{Should i have a figure showing the basis functions?} \hogg{yes}
